Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-26 WSSSPE3/report/wg_appendix_software_credit.tex
Appendix J:  A handful of changes.

The following section summarizes the working group's discussion based on
contributions prior to the meeting~\cite{WSSSPE3-SC-github-issues} and the
collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
Please refer to the original sources for the unedited discussions.

@@

\item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns
contriponents (contributors and components) various weights depending on their
level of importance; and

@@

\end{itemize}
However, as of this writing, only the Project CRediT
roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship
Badges~\cite{Mozilla_badges} have been implemented in published papers, though
both of these only provide a single ``Software'' or ``Computation'' category
associated with software. In addition, neither of these options allow for the
citation of software itself, but only provide an author contribution related to
software. The discussion quickly focused on transitive credit as a more
quantitative measure of allocating credit to both authors and software, although
there were some concerns about authors overestimating their own contributions
compared to prior work.

@@

The group also agreed on the fact that additional data was required to support
the assertion that software was not being sufficiently cited in the literature.
In particular, this issue seemed to be field-dependent. For example, as shown by
study of Howison and Bullard~\cite{Howison2015}, in the field of biology the
most-cited papers appear to be those describing scientific software. However,
this may not---and likely is not---the case in other fields, nor is it clear
whether developers of scientific software, even in the case of the biology
field, are receiving sufficient credit for their efforts.

During the first day of WSSSPE3 breakout sessions, the group discussed the
Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
scientific software. That system assigns unique Digital Object Identifiers
(DOIs)---the same system used for scientific publications---to all content
(e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of this system is to track rights and credit for contributors
to entertainment works in order to distribute revenues---similar to the proposed
transitive credit concept.

@@

On the second day of WSSSPE3, the working group decided to put aside the
taxonomy of contributions and focus on software citations to ensure developers
receive credit (regardless of contribution). Eventually, once standardizing
software citations, the goal would be to return to establishing different
roles\slash contributions for this credit. Following this decision, the group
identified two necessary actions to move forward:
\begin{enumerate}

@@

At a minimum, the metadata needed for software citation (aside from the DOI for
the software itself) includes:
\begin{itemize}
    \item Name of software,
    \item DOI for software,
    \item Contributors, in the form of names and ORCIDs,
    \item Software dependencies, in the form of DOIs, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file, e.g., as part of
the GitHub repository. The group also discussed similar efforts such as
Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an
attempt to codify minimal metadata schemes in JSON and XML for scientific
software and code, and implementing transitive credit via
JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
would be stored for closed-source software.

As one mechanism for constructing accurate contributor lists from existing
project contributors, the group discussed associating GitHub accounts---as well
as accounts on Bitbucket, CodePlex, and other repositories for open-source
scientific software---accounts with ORCID. However, a (quick) response from
GitHub (via Arfon Smith) indicated that this might not be possible in the near
future: ``GitHub doesn't have any plans to allow ORCID accounts to be associated
with GitHub user accounts.''

@@

Finally, although a discrete task from software citations themselves,
significant discussion focused on ensuring software citations are indexed in the
same manner as publications, allowing the construction of a corresponding
software citation network. Currently, software releases can receive citable DOIs
via Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these
citations are not processed by indexers such as Web of Science, Scopus, or
Google Scholar. Thus, either in parallel or following the primary task, the
group will need to reach out to these organizations Initial conversations with
Elsevier\slash Scopus (via Michael Taylor) during WSSSPE3 clarified that Scopus
is not yet DataCite DOI aware, and also does not yet have an internal identifier
for software or data (but needs\slash plans to add this support). Taylor said
they prefer a ``software article'' with the usual article metadata (e.g.,
authors, citations), and mentioned Zenodo as an example -- this proposal seemed
to align with group discussions. Taylor also mentioned another benefit of the
software and associated DOI on GitHub: in addition to a citation, could obtain
statistics on usage/downloads/forks, which happens to be what
Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

@@

Modern scientific research relies considerably on software, and ensuring that
developers of such scientific software receive credit for their efforts will
encourage additional creation and maintenance. \katznote{the prev sentence is
good, but feels a bit odd in the middle of the appendix, rather than at the
beginning} There currently is no standard mechanism for citing software or
receiving credit for software (akin to publications/citations). Software is
eligible for DOI assignment, but DOI metadata fields are not well tuned or
standardized for software (vs. publications). Some software providers apply for
DOIs but it is still not widely adopted. Also, there is no mechanism to cite
software dependencies within software.

@@

\item Hold virtual meeting to determine group members responsible\slash willing
to work on the following tasks, to be organized within one month of the workshop

\item Compile best practices of software citation across multiple disciplines,
including journals and communities of interest\slash practice in the research
world, to begin by December 2015

\item Compile examples of including other products in promotion and tenure
dossier, to begin by December 2015

\item Draft the Software Citation Principles document (including citation
metadata file), by April 2016

\item Publish\slash release the Software Citation Principles document, by August
2016

\item Reach out to journals, publishers, teachers\slash educators, indexers, and
professional societies---likely through meetings with key groups, to begin by
September 2016

@@

Following the meeting, Kyle Niemeyer formally requested the merging of software
citation activities with FORCE11-SCWG. Within a month of the meeting, Kyle will
organize a virtual meeting of the group and lead the division of
responsibilities for compiling existing practices of software citation and
including software\slash products in promotion and tenure dossiers. Building off
of these efforts, the next major milestone is drafting the Software Citation
Principles document in collaboration with the SCWG, targeted for April 2016.
While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the
efforts of that group towards the Software Citation Principles document, Kyle
will help coordinate contributions from the WSSSPE group members.

>>>>

The following section summarizes the working group's discussion based on
contributions prior to the meeting~\cite{WSSSPE3-SC-github-issues} and the
collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
Please refer to the original sources for the unedited discussions if necessary.

@@

\item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns
contriponents (contributors and components) various weights depending on their
level of importance;

@@

\end{itemize}
However, as of this writing, only Project CRediT
roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship
Badges~\cite{Mozilla_badges} have been implemented for published papers, and
both of these only provide a single ``Software'' or ``Computation'' category
associated with software. In addition, neither of these options allows for the
citation of software itself, but only provide an author contribution related to
software. The discussion quickly focused on transitive credit as a more
quantitative measure of allocating credit to both authors and software, although
there were some concerns about authors overestimating their own contributions
compared to prior work.

@@

The group also agreed on the fact that additional data was required to support
the assertion that software was not being sufficiently cited in the literature.
In particular, this issue seemed to be field-dependent. For example, as shown by
a study of Howison and Bullard~\cite{Howison2015}, in the field of biology, the
most-cited papers appear to be those describing scientific software. However,
this may not---and likely is not---the case in other fields, nor is it clear
whether developers of scientific software, even in the case of the biology
field, are receiving sufficient credit for their efforts.

During the first day of WSSSPE3 breakout sessions, the group discussed the
Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
scientific software. That system assigns unique Digital Object Identifiers
(DOIs)---the same system used for scientific publications---to all content
(e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of this system is to track rights and credits for contributors
to entertainment works in order to distribute revenues---similar to the proposed
transitive credit concept.

@@

On the second day of WSSSPE3, the working group decided to put aside the
taxonomy of contributions and focus on software citations to ensure developers
receive credit (regardless of contribution). Eventually, once 
software citations are standardized, the goal would be to return to establishing different
roles\slash contributions for this credit. Following this decision, the group
identified two necessary actions to move forward:
\begin{enumerate}

@@

At a minimum, the metadata needed for software citation %(aside from the DOI for
%the software itself) \choinote{DOI is in the list below}
includes:
\begin{itemize}
    \item Name of software,
    \item DOI for software,
    \item Contributors, in the form of names and ORCIDs,
    \item Software dependencies, in the form of DOIs, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file, e.g., as part of
the GitHub repository. The group also discussed similar efforts such as
Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an
attempt to codify minimal metadata schemes in JSON and XML for scientific
software and code, and implementing transitive credit via
JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
would be stored for closed-source software.

As one mechanism for constructing accurate contributor lists from existing
project contributors, the group discussed associating GitHub accounts---as well
as accounts on Bitbucket, CodePlex, and other repositories for open-source
scientific software---with ORCID accounts. However, a (quick) response from
GitHub (via Arfon Smith) indicated that this might not be possible in the near
future: ``GitHub doesn't have any plans to allow ORCID accounts to be associated
with GitHub user accounts.''

@@

Finally, although a discrete task from software citations themselves,
significant discussion focused on ensuring software citations are indexed in the
same manner as publications, allowing the construction of a corresponding
software citation network. Currently, software releases can receive citable DOIs
via Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these
citations are not processed by indexers such as Web of Science, Scopus, or
Google Scholar. Thus, either in parallel or following the primary task, the
group will need to reach out to these organizations. Initial conversations with
Elsevier\slash Scopus via Michael Taylor during WSSSPE3 clarified that Scopus
is not yet DataCite DOI aware, and also does not yet have an internal identifier
for software or data (but needs\slash plans to add this support). Taylor said
they prefer a ``software article'' with the usual article metadata (e.g.,
authors, citations), and mentioned Zenodo as an example -- this proposal seemed
to align with our group discussions. Taylor also mentioned another benefit of the
software and associated DOI on GitHub: in addition to a citation, one could obtain
statistics on usage/downloads/forks, which happens to be what
Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

@@

Modern scientific research relies considerably on software, and ensuring that
developers of such scientific software receive credit for their efforts will
encourage additional creation and maintenance. \katznote{the prev sentence is
good, but feels a bit odd in the middle of the appendix, rather than at the
beginning} There currently is no standard mechanism for citing software or
receiving credit for software (akin to publications/citations). Software is
eligible for DOI assignment, but DOI metadata fields are not well tuned or
standardized for software (vs.\ publications). Some software providers apply for
DOIs, but the identifiers are still not widely adopted. Also, there is no mechanism to cite
software dependencies within software.

@@

\item Hold virtual meeting to determine group members responsible\slash willing
to work on the following tasks, to be organized within one month of the workshop.

\item Compile best practices of software citation across multiple disciplines,
including journals and communities of interest\slash practice in the research
world, to begin by December 2015.

\item Compile examples of including other products in promotion and tenure
dossier, to begin by December 2015.

\item Draft the Software Citation Principles document (including citation
metadata file), by April 2016.

\item Publish\slash release the Software Citation Principles document, by August
2016.

\item Reach out to journals, publishers, teachers\slash educators, indexers, and
professional societies---likely through meetings with key groups, to begin by
September 2016.

@@

Following the meeting, Kyle Niemeyer formally requested the merging of software
citation activities with FORCE11-SCWG. Within a month of the meeting, Niemeyer will
organize a virtual meeting of the group and lead the division of
responsibilities for compiling existing practices of software citation and
including software\slash products in promotion and tenure dossiers. Building off
of these efforts, the next major milestone is drafting the Software Citation
Principles document in collaboration with the SCWG, targeted for April 2016.
While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the
efforts of that group towards the Software Citation Principles document, Kyle
will help coordinate contributions from the WSSSPE group members.
